---
title: "Roy Model"
output:
  pdf_document: default
  html_document: default
---



```{r setup}
# We will load the tidyverse, a commonly used set of R libraries
# Find more information here: https://www.tidyverse.org/packages/
library('tidyverse')
library('MASS')
```

## Roy Model Simulation 1

### Generate Random Data

Economist earnings are normally distributed with mean = 60k and stdev = 10k, and Accountant earnings are normally distributed with mean = 65k and st dev = 5k. This generation will produce data that pairs draws from both distribution, i.e. we are observing for every theoretical individual in our sample both their potential earnings as an accountant and their potential earnings as an economist. 

```{r data}
set.seed(10042018)

samples <- 1000000  # Sample size or size of population
cor <- 0.9 # correlation coefficient

# Generate correlated std normal random sample using the mvnorm function from MASS package
data <- mvrnorm(n=samples, mu=c(0, 0), Sigma=matrix(c(1, cor, cor, 1), nrow=2), empirical=FALSE)

# Convert to dataframe
df1 <- as_data_frame(data)

# Make the dataframe easier to use 
df1 <- df1 %>% dplyr::rename(accnt = V1, econ = V2) # rename columns

# Change distributions from std normal to those specified in lecture:
# Accounting ~ N(65000, 5000)
# Economics ~ N(60000, 10000)
mu_econ <- 60000
sigma_econ <- 10000
mu_accnt <- 65000
sigma_accnt <- 5000

df1 <- df1 %>% mutate(
        accnt = accnt*sigma_accnt + mu_accnt, # update accounting variable
        econ = econ*sigma_econ + mu_econ # update econ variable
)
```

### Sanity Checks

Let's run a few checks to make sure data looks reasonable. First we'll use the `head` and `tail` commands to get a look at the data. Then, we will make sure that the correlation between the accoutant and economist distributions is 0.86, the value we set when we generated the data, using the `cor` command. Finally, we will run the `summarise` command to see the mean, min, max, and quartiles of each of the distributions.

```{r sanity}
# Look at the first and last 6 rows of the dataframe
head(df1)
tail(df1)

# Compare the correlation we set to the correlation we calculate
check <- round(cor(df1$accnt, df1$econ)) == round(cor)
print(ifelse(check, "The correlations are the same!", "Oops, the correlations are not the same."))

# Generate summary statistics 
print("Economists Summary Stats")
summary(df1$econ)
print("Accountant Summary Stats")
summary(df1$accnt)

### Picking a Career

#Our assumption is that every person will choose to be an accountant or an 
#economist based on what will maximize their salary. We will assign job labels 
#strictly based on where an individual will earn more, 
#i.e. Y<sub>i</sub> = max(Y<sub>e,i</sub>, Y<sub>a,i</sub>)
#```{r jobs}
# Assign job labels
df1 <- df1 %>% mutate(job = ifelse(econ > accnt, "econ", "accnt"))

# Take a look at the change to the dataframe using head
head(df1)
```

### Results

```{r results}
# Make results dataframe
results <- df1 %>% 
            group_by(job) %>% # Group all of the rows with the same "job" together
            summarise('Economist Earnings' = mean(econ), 'Accountant Earnings' = mean(accnt),
                      
            n=n()) %>% # Calculate means and counts for economicts and accountants
            mutate(job = c("Accountant", "Economist")) %>% # add labels
            t() # transpose

colnames(results) <- c('Accountant', 'Economist')




results <- as.data.frame(results) %>%
            slice(2:4) %>%
            mutate(
              x= c('Economist Earnings', 'Accountant Earnings', 'n')
            ) %>%
            dplyr::select(x, Accountant, Economist)

results
```







## Roy Model Simulation 2

### Generate Random Data

Example: Economist earnings are normally distributed with mean = 60k and stdev = 10k, and Accountant earnings are normally distributed with mean = 65k and st dev = 5k. This generation will produce data that pairs draws from both distribution, i.e. we are observing for every theoretical individual in our sample both their potential earnings as an accountant and their potential earnings as an economist. 

```{r data2}
set.seed(02101870)


samples <- 1000000  # Sample size or size of population
cor <- 0.9 # correlation coefficient

# Generate correlated std normal random sample using the mvnorm function from MASS package
data <- mvrnorm(n=samples, mu=c(0, 0), Sigma=matrix(c(1, cor, cor, 1), nrow=2), empirical=FALSE)

# Convert to dataframe
df1 <- as_data_frame(data)

# Make the dataframe easier to use 
df1 <- df1 %>% dplyr::rename(accnt = V1, econ = V2) # rename columns

# Change distributions from std normal to those specified in lecture:
# Accounting ~ N(65000, 5000)
# Economics ~ N(60000, 10000)
mu_econ <- 60000
sigma_econ <- 10000
mu_accnt <- 65000
sigma_accnt <- 5000

df1 <- df1 %>% mutate(
        accnt = accnt*sigma_accnt + mu_accnt, # update accounting variable
        econ = econ*sigma_econ + mu_econ # update econ variable
)
```

### Sanity Checks

Let's run a few checks to make sure data looks reasonable. First we'll use the `head` and `tail` commands to get a look at the data. Then, we will make sure that the correlation between the accoutant and economist distributions is 0.86, the value we set when we generated the data, using the `cor` command. Finally, we will run the `summarise` command to see the mean, min, max, and quartiles of each of the distributions.

```{r sanity2}
# Look at the first and last 6 rows of the dataframe
head(df1)
tail(df1)

# Compare the correlation we set to the correlation we calculate
check <- round(cor(df1$accnt, df1$econ)) == round(cor)
print(ifelse(check, "The correlations are the same!", "Oops, the correlations are not the same."))

# Generate summary statistics 
print("Economists Summary Stats")
summary(df1$econ)
print("Accountant Summary Stats")
summary(df1$accnt)

### Picking a Career

#Our assumption is that every person will choose to be an accountant or an 
#economist based on what will maximize their salary. We will assign job labels 
#strictly based on where an individual will earn more, 
#i.e. Y<sub>i</sub> = max(Y<sub>e,i</sub>, Y<sub>a,i</sub>)
#```{r jobs}
# Assign job labels
df1 <- df1 %>% mutate(job = ifelse(econ > accnt, "econ", "accnt"))

# Take a look at the change to the dataframe using head
head(df1)
```

### Results

```{r results2}
# Make results dataframe
results <- df1 %>% 
            group_by(job) %>% # Group all of the rows with the same "job" together
            summarise('Economist Earnings' = mean(econ), 'Accountant Earnings' = mean(accnt),
            n=n()) %>% # Calculate means and counts for economicts and accountants
            mutate(job = c("Accountant", "Economist")) %>% # add labels
            t() # transpose

colnames(results) <- c('Accountant', 'Economist')




results <- as.data.frame(results) %>%
            slice(2:4) %>%
            mutate(
              x= c('Economist Earnings', 'Accountant Earnings', 'n')
            ) %>%
            dplyr::select(x, Accountant, Economist)

results
```






## Roy Model Simulation 3


### Generate Random Data

Example : Economist earnings are normally distributed with mean = 60k and stdev = 10k, and Accountant earnings are normally distributed with mean = 65k and st dev = 5k. This generation will produce data that pairs draws from both distribution, i.e. we are observing for every theoretical individual in our sample both their potential earnings as an accountant and their potential earnings as an economist. 

```{r data3}
set.seed(10042018)



samples <- 1000000  # Sample size or size of population
cor <- 0.9 # correlation coefficient

# Generate correlated std normal random sample using the mvnorm function from MASS package
data <- mvrnorm(n=samples, mu=c(0, 0), Sigma=matrix(c(1, cor, cor, 1), nrow=2), empirical=FALSE)

# Convert to dataframe
df1 <- as_data_frame(data)

# Make the dataframe easier to use 
df1 <- df1 %>% dplyr::rename(accnt = V1, econ = V2) # rename columns

# Change distributions from std normal to those specified in lecture:
# Accounting ~ N(65000, 5000)
# Economics ~ N(60000, 10000)
mu_econ <- 60000
sigma_econ <- 10000

mu_accnt <- 65000
sigma_accnt <- 5000

df1 <- df1 %>% mutate(
        accnt = accnt*sigma_accnt + mu_accnt, # update accounting variable
        econ = econ*sigma_econ + mu_econ # update econ variable
)
```

### Sanity Checks

Let's run a few checks to make sure data looks reasonable. First we'll use the `head` and `tail` commands to get a look at the data. Then, we will make sure that the correlation between the accoutant and economist distributions is 0.86, the value we set when we generated the data, using the `cor` command. Finally, we will run the `summarise` command to see the mean, min, max, and quartiles of each of the distributions.

```{r sanity3}
# Look at the first and last 6 rows of the dataframe
head(df1)
tail(df1)

# Compare the correlation we set to the correlation we calculate
check <- round(cor(df1$accnt, df1$econ)) == round(cor)
print(ifelse(check, "The correlations are the same!", "Oops, the correlations are not the same."))

# Generate summary statistics 
print("Economists Summary Stats")
summary(df1$econ)
print("Accountant Summary Stats")
summary(df1$accnt)

### Picking a Career

# Our assumption is that every person will choose to be an accountant or an 
# economist based on what will maximize their salary. We will assign job labels 
# strictly based on where an # individual will earn more, 
# i.e. Y<sub>i</sub> = max(Y<sub>e,i</sub>, Y<sub>a,i</sub>)
#```{r jobs}
# Assign job labels
df1 <- df1 %>% mutate(job = ifelse(econ > accnt, "econ", "accnt"))

# Take a look at the change to the dataframe using head
head(df1)
```

### Results

```{r results3}
# Make results dataframe
results <- df1 %>% 
            group_by(job) %>% # Group all of the rows with the same "job" together
            #summarise('Economist Earnings' = mean(econ), 'Accountant Earnings' = mean(accnt),
                      
       
              #Changing the code to calculate and display standard deviation
              summarise('Economist Earnings' = sd(econ), 'Accountant Earnings' = sd(accnt),

                      n=n()) %>% # Calculate sd and counts for economicts and accountants
            mutate(job = c("Accountant", "Economist")) %>% # add labels
            t() # transpose

colnames(results) <- c('Accountant', 'Economist')




results <- as.data.frame(results) %>%
            slice(2:4) %>%
            mutate(
              x= c('Economist Earnings', 'Accountant Earnings', 'n')
            ) %>%
            dplyr::select(x, Accountant, Economist)

results
```



The standard normal sample size is 1000000
It is observed that out of the sample size for economists only 198691 have become economists and hence the standard deviation we observed (5709.471) is significantly less compared to the conditional standard deviation of 100000
The same goes for accountants, out of the sample size for accountants its observed that 801309 have become accountants, hence the standard deviation we observed
(4496.64) is less compared to the conditional standard deviation set of 5000



## Roy Model Simulation 4


### Generate Random Data

Example: Economist earnings are normally distributed with mean = 60k and stdev = 10k, and Accountant earnings are normally distributed with mean = 65k and st dev = 5k. This generation will produce data that pairs draws from both distribution, i.e. we are observing for every theoretical individual in our sample both their potential earnings as an accountant and their potential earnings as an economist. 

```{r data4}
set.seed(10042018)


samples <- 1000000  # Sample size or size of population
cor <- 0.9 # correlation coefficient

# Generate correlated std normal random sample using the mvnorm function from MASS package
data <- mvrnorm(n=samples, mu=c(0, 0), Sigma=matrix(c(1, cor, cor, 1), nrow=2), empirical=FALSE)

# Convert to dataframe
df1 <- as_data_frame(data)

# Make the dataframe easier to use 
df1 <- df1 %>% dplyr::rename(accnt = V1, econ = V2) # rename columns

# Change distributions from std normal to those specified in lecture:
# Accounting ~ N(65000, 5000)
# Economics ~ N(60000, 10000)
mu_econ <- 60000

  sigma_econ <- 12000

mu_accnt <- 65000
sigma_accnt <- 5000

df1 <- df1 %>% mutate(
        accnt = accnt*sigma_accnt + mu_accnt, # update accounting variable
        econ = econ*sigma_econ + mu_econ # update econ variable
)
```

### Sanity Checks

Let's run a few checks to make sure data looks reasonable. First we'll use the `head` and `tail` commands to get a look at the data. Then, we will make sure that the correlation between the accoutant and economist distributions is 0.86, the value we set when we generated the data, using the `cor` command. Finally, we will run the `summarise` command to see the mean, min, max, and quartiles of each of the distributions.

```{r sanity4}
# Look at the first and last 6 rows of the dataframe
head(df1)
tail(df1)

# Compare the correlation we set to the correlation we calculate
check <- round(cor(df1$accnt, df1$econ)) == round(cor)
print(ifelse(check, "The correlations are the same!", "Oops, the correlations are not the same."))

# Generate summary statistics 
print("Economists Summary Stats")
summary(df1$econ)
print("Accountant Summary Stats")
summary(df1$accnt)

### Picking a Career

#Our assumption is that every person will choose to be an accountant or an 
#economist based on what will maximize their salary. We will assign job labels 
#strictly based on where an individual will earn more, 
#i.e. Y<sub>i</sub> = max(Y<sub>e,i</sub>, Y<sub>a,i</sub>)
#```{r jobs}
# Assign job labels
df1 <- df1 %>% mutate(job = ifelse(econ > accnt, "econ", "accnt"))

# Take a look at the change to the dataframe using head
head(df1)
```

### Results

```{r results4}
# Make results dataframe
results <- df1 %>% 
            group_by(job) %>% # Group all of the rows with the same "job" together
            summarise('Economist Earnings' = mean(econ), 'Accountant Earnings' = mean(accnt),
                      

                      n=n()) %>% # Calculate means and counts for economicts and accountants
            mutate(job = c("Accountant", "Economist")) %>% # add labels
            t() # transpose

colnames(results) <- c('Accountant', 'Economist')




results <- as.data.frame(results) %>%
            slice(2:4) %>%
            mutate(
              x= c('Economist Earnings', 'Accountant Earnings', 'n')
            ) %>%
            dplyr::select(x, Accountant, Economist)

results
```


The data from this simulation  shows an increase in total economists and decrease in accountants. As the standard deviation of economist earnings is increased from $10000 to $12000, the economist earnings of economist have increased and accountants with economist earnings have reduced.
This is due to the flattening of the curve that happens because of the increase in standard deviation causing the occurrences of high economist earnings has resulted accountants becoming economists, which means increase in total economists and decrease in total accountants. Due to the same, the mean of economists with economist earnings increases (Mean changed from 73066.65 to 74370.56 ) whereas the mean of accountants with accountant earn
ings has reduced(Mean changed from 63822.98 to 63371.95). One another observation is the increase in people with economists earnings (198340 to 260672) and reduction in number of people with accountants earnings (801660 to 739328)

